Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-09-08 chapters/chapter06.tex

New analyses will also be made attainable, with a possibility to produce spatial statistics such as wavefront \citep{Pybus2012}, for visualising emerging epidemics.
The new release of SPREAD will also have a Command Line Interface that will make it easier to pipeline large repeatable analysis, produce scripts or call SPREAD from interpretable languages like R \citep{RCran} or Python.

@@

With the exclusion of empirical models of evolution, most phylogenetic methods are built on mathematical theories that predict outcomes conditional on some specified assumptions.
How well these models fit the real data, and how robust they are to the violations of these assumptions remains an open question.
Furthermore, the complex biological systems being the topic of phylogenetic studies undergo many processes, which interact and are difficult to accurately model or distinguish from the background noise in the data.
Rapidly evolving viruses, which are the main interest of the research presented in this thesis, are subject to mutation, natural selection and spatial diffusion.
Therefore for most of the molecular epidemiology data-sets the true underlying evolutionary process that generated them remains unknown.
All of these factors stress the need for development of flexible simulation software, that can match the diversity, complexity and sheer amount of the real data that is currently being sampled.

% adding more models
\subsection{Model availiability}

Our phylogenetic simulation software $\pi$BUSS presented in Chapter \ref{chap:pibuss} allows for fabricating evolution under a vast array of coalescent, amino-acid, nucleotide and codon substitution models, as well as diffusion models, combined with various molecular clock models.
In addition it offers an ability to apply different models across arbitrary partitioning schemes.
In future releases of $\pi$BUSS the collection of available models will be extended even further.
The development effort will be aimed at matching the model richness available for inference in BEAST, and supplying every method with its simulation counterpart in $\pi$BUSS.

@@

% gui for more complex models
The models' specification will continue to be available via the user-friendly Graphical User Interface (GUI), which allows to setup a simulation by choosing models from drop-down menus, partitioning data in tables and parsing parameters from text fields in an intuitive fashion.
Time-heterogenous models, for which evolutionary parameters change both across different lineages (see Chapter \ref{chap:dpp}) as well as models of pan-lineage change (see Chapter \ref{chap:epoch}) will have dedicated panels, where users can formulate them without the need to manually edit XML files.
% beauti integration
In addition to it's Command Line Interface for scripting purposes and XML parsers available via BEAST's core implementation, $\pi$BUSS is in a sense similar to the BEAUti software package, which helps BEAST users in setting up their analysis through a user-friendly GUI.
We will follow along that path, by providing even closer integration between the two programs and facilitating generation of joint simulation-analysis XML documents in which fabricated data is instantly passed to BEASTs XML parsers for inference.

@@

%GB: this first sentence is unclear, please rewrite it.
In the field of molecular biology insertions and deletions (indels) exist on the other end of spectrum of mutations then substitution events, which we discussed in details in Subsection \ref{sub:subst_models}.
Whereas substitutions are point mutations which replace one nucleotide with another, indels proceed by either inserting or deleting nucleotides in a sequence, thus altering its length. 
Indel mutations are well known to have a significant impact on processes of molecular evolution \citep{Fletcher2009}.
Therefore without models of insertion/deletion sequence simulation is somewhat limited.
Future releases of $\pi$BUSS will implement that option, starting with the stochastic Dollo model \citep{LeQuesne1974}.

@@

The ever growing availability of molecular sequences produced by high-throughput sequencing yields data which is difficult to store, let alone analyse.
This problem is dubbed the biggest challenge for the field of molecular virology. % citation

Furthermore due to hardware limitations and problems with heat dissipation we are approaching the limit of how many transistors can be put into integrated circuits.
%GB: need citation for Moore's law?
An observation which came to be known as Moore's law predicted the trend in which chip performance should double roughly every 18 months.
However because of these problems serial computations on next generations of CPUs no longer yield the same increases in speed and performance.
% citation
If we want to capitalize on vast amounts of data that are available, future development of phylogenetic software should target high performance parallel devices such as GPUs \citep{Nickolls2008} or Field-programmable gate arrays (FPGA, \citet{Kuon2008}).

@@

Interestingly, realistic models of evolution also bring new computational challenges with them.
Models with large state spaces, such as codon models discussed in Chapter \ref{chap:dpp} are an example of such a challenge.
Fortunately recent advances in computing on GPUs \citep{Ayres2012} provide opportunities to accelerate phylogenetic inference.
However we believe that more research will be needed to mitigate the imminent computational burdens that come with novel modelling approaches and high-throughput sequencing.

% visualisations
\subsection{Big data}

The advent of next generation sequencing data is expected to allow us to generate complete genome sequences swiftly and a fraction of todays cost.
This, combined with high-performance computing, will allow us to analyse the data in almost real time, and by coupling molecular sequence data with geographical information, to trace the diffusion of emerging epidemics and instantly identify potential hosts and sinks as well as other factors contributing to viral spread.

>>>>

New analyses will also be made attainable, with a possibility to produce spatial statistics such as wavefront \citep{Pybus2012}, for visualising emerging epidemics.
%PL: provide a bit more background on these spatial stats: wavefront, dispersal rate, diffusion coefficient, how they are drawn from continuous diffusion reconstructions and what they can do in terms of comparative analyses and hypothesis testing..
The new release of SPREAD will also have a Command Line Interface that will make it easier to pipeline large repeatable analysis, produce scripts or call SPREAD from interpretable languages like R \citep{RCran} or Python.

%PL: you could also mention compatibility with BEAST2 and take this as an opportunity to describe the current situation of divergence in the two beast versions and parallel development for both.. the reader might be interested in this.

@@

Except for empirical models of evolution, most phylogenetic methods are built on mathematical theories that predict outcomes conditional on a set of assumptions.
How well these models fit the real data, and how robust they are to the violations of these assumptions remains an open question.
Furthermore, complex biological systems targeted phylogenetic studies undergo many processes, which interact and are difficult to accurately model or distinguish from the background noise in the data.
Rapidly evolving viruses, which are the main interest of the research presented in this thesis, are subject to mutation, natural selection and spatial diffusion.
Therefore, for most of the molecular epidemiology data-sets, the true underlying evolutionary process that generated them remains unknown or particular aspects of the process remain difficult to test.
All of these factors stress the need for development of flexible simulation software, that can match the diversity, complexity and sheer amount of the real data that is currently being sampled.

% adding more models
\subsection{Model availability}

Our phylogenetic simulation software $\pi$BUSS presented in Chapter \ref{chap:pibuss} allows for fabricating evolution under a variety of coalescent, amino-acid, nucleotide and codon substitution models, as well as diffusion models, combined with various molecular clock models.
In addition it offers an ability to apply different models across arbitrary partitioning schemes. %PL: also epochs?
In future releases of $\pi$BUSS the collection of available models and data types will be extended even further.
The development effort will be aimed at matching the model richness available for inference in BEAST, and supplying every method with its simulation counterpart in $\pi$BUSS.
%PL: mention more explicitly what to add meat to this: e.g., flexible coalescent priors (skyline/ride/grid), the codon models extension you are aiming at developing, and importantly continuous trait data. We are currently moving into the field of comparative biology with the continuous framework, so there is a need to be able to simulate multivariate trait data, and this under several models such as Brownian, relaxed random walks, Ornstein-Uhlenbeck or drift models... this could be separate item almost like index-evolution

@@

% gui for more complex models
The models' specification will continue to be available via the user-friendly Graphical User Interface (GUI), which allows to setup a simulation by choosing models from drop-down menus, partitioning data in tables and parsing parameters from text fields in an intuitive fashion.
Time-heterogenous models, for which evolutionary parameters change both across different lineages (see Chapter \ref{chap:dpp}) as well as models of pan-lineage change (see Chapter \ref{chap:epoch}) will have dedicated panels, where users can formulate them without the need to manually edit XML files.
% beauti integration
In addition to it's Command Line Interface for scripting purposes and XML parsers available via BEAST's core implementation, $\pi$BUSS is in a sense similar to the BEAUti software package, which helps BEAST users in setting up their analysis through a user-friendly GUI.
We will proceed along that path, by providing even closer integration between the two programs and facilitating generation of joint simulation-analysis XML documents in which fabricated data is instantly passed to BEASTs XML parsers for inference.

@@

%GB: this first sentence is unclear, please rewrite it.
In the field of molecular biology, insertions and deletions (indels) exist on the other end of spectrum of mutations then substitution events, which we discussed in details in Subsection \ref{sub:subst_models}.
Whereas substitutions are point mutations which replace one nucleotide with another, indels proceed by either inserting or deleting nucleotides in a sequence, thus altering its length. 
Indel mutations are well known to have a significant impact on processes of molecular evolution \citep{Fletcher2009}. %PL: perhaps mention that they require different models and assumptions such as reversibility are not realistic for such processes
Therefore without models of insertion/deletion sequence simulation is somewhat limited.
Future releases of $\pi$BUSS will implement that option, starting with the stochastic Dollo model \citep{LeQuesne1974}.
%PL: you could mention why indel processes are useful -- I would think this is more for evaluating alignment procedures - and take this as an opportunity to talk about statistical alignment and joint inference of alignment and phylogeny (a la Marc's Baliphy), and that it would be great to get this in BEAST, but mention the computational hindrances. I think one of Marc's disciples also showed how to integrate such approaches in codon model approaches to identify positive selection: http://people.duke.edu/~br51/branch-site-article.pdf

@@

The ever growing availability of molecular sequences produced by high-throughput sequencing yields data which is difficult to store, let alone to analyze.
This problem is dubbed the biggest challenge for the field of molecular virology. % citation %PL: probably not just virology but molecular evolution in general?

%PL: the next sentence imemediately jumps into specifics, maybe introduce HPC and a bit of its history/strategies before coming to the limitations.
Furthermore due to hardware limitations and problems with heat dissipation we are approaching the limit of how many transistors can be put into integrated circuits.
%GB: need citation for Moore's law?
An observation which came to be known as Moore's law predicted the trend in which chip performance should double roughly every 18 months.
However because of these problems serial computations on next generations of CPUs no longer yield the same increases in speed and performance.
% citation
If we want to capitalize on vast amounts of data that are available, future development of phylogenetic software should target high performance parallel devices such as GPUs \citep{Nickolls2008} or Field-programmable gate arrays (FPGA, \citet{Kuon2008}).
%PL: here's an opportunity to talk a bit more about these things (I for one don't really know what FPGA are). GPUs are really good for particular tasks, but surely they are not the holy grail of HPC either, so talk about their advantages, drawbacks and some future perspectives... Maybe also talk about BEAGLE, its current CUDA dependency, and what the prospects are here. This is stuff that David Posada might be really interested in.

@@

Interestingly, realistic models of evolution also bring about new computational challenges.
Models with large state spaces, such as codon models discussed in Chapter \ref{chap:dpp} are an example of such a challenge.
Fortunately recent advances in computing on GPUs \citep{Ayres2012} provide opportunities to accelerate phylogenetic inference.
However we believe that more research will be needed to mitigate the imminent computational burdens that come with novel modeling approaches and high-throughput sequencing. 
%PL: this paragraph doesn't say much. Perhaps when you put the lineage codon models with DP process priors in the discussion, you can extend on this?

% visualisations
\subsection{Big data}
%PL this is OK as concluding paragraph in another section, but it doesn't warrant a separate subsection as 'Big data.
The advent of next generation sequencing data has lead to the generation of complete genome sequences in short time and at relatively low costs.
This, combined with high-performance computing, will allow us to analyse the data in almost real time, and by coupling molecular sequence data with geographical information, to trace the diffusion of emerging epidemics and instantly identify potential hosts and sinks as well as other factors contributing to viral spread.
